Motional Quantum Ground-State Cooling Outside the Lamb-Dicke Regime

### 2017-06-28 paper.tex
Add description of the sequence

The Raman sideband cooling we use to achieve the high ground state preparation fidelity
consists of multiple cycles of laser pulses to manipulate the internal and
motional states of the atoms.
Figure \ref{f-setup}A shows the schematic of the energy levels and the cooling sequence
in our setup.
Each cooling cycle starts with the sodium atom in the $|F=2, m_F=-2\rangle$
ground electronic state and a certain vibration state $n$.
In the first step, a Raman pulse drives a transition to the motional state $n-\Delta n$
to reduce the motional energy while also changing the internal state to $|F=1, m_F=-1\rangle$.
In the second step, which finishes the cooling cycle,
an optical pumping pulse bring the atom back to the $|F=2, m_F=-2\rangle$ state to take away
the entropy via the spontaneously emitted photon.
The second step could also change the motional state of the atoms and result in heating.
The possibility for this to happen for an atom in the motional level $n$
is approximately proportional to the effective Lamb-Dicke (LD) parameter
$\eta^{OP}_{\eff}=\sqrt{2n+1}\eta^{OP}$ where $\eta^{OP}=k^{OP}\sqrt{\hbar/2m\omega}$
is the LD parameter for optical pumping (OP), $k^{OP}$ is the wavenumber of the OP light, $m$ is the mass of the atom, and $\hbar$ is the reduced Planck constant.

@@

In order to perform Raman sideband cooling efficiently and minimize
the heating during the optical pumping, we need to have a low initial temperature and
a small LD parameter. However, since the LD parameter $\eta\propto 1/(\sqrt[4]{m}\lambda)$ where $m$ is the mass of the atom and $\lambda$
is the wavelength of the atomic transition,
it is larger for sodium at a given trap depth due to its low mass and short D line wavelength.
With 45 mW of power in the trap, we measure trapping frequencies of
$\{\omega_1,\omega_2,\omega_3\}/2\pi = \{69(1), 430(4), 590(5)\}\ \text{\text{kHz}}$
which corresponds to optical pumping LD parameters of
$\{\eta^{OP}_1,\eta^{OP}_2,\eta^{OP}_3\} = \{0.602(5), 0.241(2), 0.206(1)\}$
Moreover, due to the small hyperfine splitting in the $3^2P_{3/2}$ manifold,
the sub-Doppler cooling in sodium is also less efficient and we start the
Raman sideband cooling with a initial temperature of 70 $\mu$K. Combined with the high $\eta$, this gives us a initial effective optical pumping LD parameters of
$\{\eta^{OP}_{1\eff},\eta^{OP}_{2\eff},\eta^{OP}_{3\eff}\} = \{4.0(1), 0.67(2), 0.50(1)\}$
As a result, there is a very high average change of motional state during the optical pumping step
(figure \ref{f-ld}B). The averaged change in motional states over the initial distribution
is $2.4$ in the axial direction.
Compare to $0.89-0.95$ in the axis with the weakest confinement in previous experiments
\cite{Thompson2013,Kaufman2012,Li2012,Liu2017} this causes greater uncertainty in the motional
state making it harder to cool.

@@

For the more tightly confined radial directions,
we begin the cooling with $\{\bar n_2, \bar n_3\}=3.4(2), 2.5(2)$.
The radial sideband spectra of the initial distribution is shown in figure \ref{f-radial}A
as orange and red lines, where we can clearly see the first order heating,
first order cooling and second order cooling sidebands.
After applying around 1000 cooling pulses cooling in all three dimensions
starting with cooling on the radial second order,
the Raman spectrum with the same parameters is shown in figure \ref{f-radial}A as
blue and green lines, where the first and second order cooling sidebands
on both axis are suppressed.
Given the absence of the second order cooling sideband,
we can estimate the ground state probability in each direction based on the height of
the first order cooling and heating sideband to be $90(2)\%$ and $94(3)\%$.

@@

The axial direction is much less confined and therefore has a much higher initial
$\bar n_1=21(2)$ and effective LD parameter of $\eta^{OP}_{eff1}\approx 4.0(2)$,
and is therefore more difficult to cool.
The red line in figure \ref{f-axial}A shows the Raman spectrum in the axial direction
before applying Raman sideband cooling; the clearly resolved sidebands
up to eighth order indicates population in highly-excited motional levels.
Therefore, we begin our cooling sequence by driving Raman transitions on the eighth order axial
cooling sideband. After applying the cooling sequence identical to the one we use to cool
the radial directions, the spectrum is shown in figure \ref{f-axial}A as the blue line.
All of the high order ($\geqslant2$) axial cooling sidebands disappear, and the first order
cooling sideband is strongly suppressed.
The ground state probability calculated from this spectrum is $92(3)\%$.
We can again use Rabi flopping on the carrier and heating sideband to verify this result
similar to the radial direction. In this case, we see very good agreement on the carrier
(figure \ref{f-axial}B) but there is additional decoherence on the axial first order
heating sideband (figure \ref{f-axial}C).
We believe this decoherence is caused by technical noise, which is the most pronounced
on this transition due to its slow Rabi frequency.
The decoherence time scale is consistent with magnetic field fluctuations of $1.5$ mG that we have measured in the lab, which produces
a Zeeman shift of around $3$ kHz.

Combining the axial and radial cooling results,
we have achieved a 3D ground state probability of $77(4)\%$.
We believe this is currently limited by off-resonant scattering from the Raman beams,
which are measured to be between $3$ to $15$ kHz for the detuning and power we use,
and the resonance shift caused by magnetic field fluctuations.
Other imperfections of the current experiement include a $5\%$ loss during the single atom
imaging and an additional $10\%$ due to the heating from the Raman beams during cooling.
We are planing to improve these by increasing the detuning of the Raman beams,
implementing better control of the magnetic field and optimize the temperature during imaging.
We expect these changes to improve the cooling performance even further.
We have shown that despite the difficulty in cooling sodium atoms,
it is possible to achieve reliable three dimensional cooling with significant ground state
probability using new cooling techniques and a optimized cooling sequence.
This opens up the possibility of cooling a large variety of systems to their motional ground
state using Raman sideband cooling that is challenging to laser cool
including molecules and exotic atoms.

% Missing:
% ?? Blackman pulse vs square pulse
% ?? Start from highest visible order
% In the text, more description
% Detail in appendix

>>>>

The experiment begins by loading the optical tweezer from a magneto-optical trap.
The tweezer is created by focusing a $700$nm beam through an objective with NA $0.55$.
With $45$mW of power, we measure a trapping frequencies of
$\{\omega_1,\omega_2,\omega_3\}/2\pi = \{69(1), 430(4), 590(5)\}\ \text{\text{kHz}}$.
The successfulness of the loading is detected by taking an image of the single atom for $1.5$ms.
The same imaging light is optimized for cooling and reduces the temperature of the single atom
to $70$ $\mu$K which corresponds to an initial motional states of
$\{\bar n_1, \bar n_2, \bar n_3\}=21(2), 3.4(2), 2.5(2)$.

The Raman sideband cooling we use to further reduce the temperature of the single atom and
achieve the high ground state preparation fidelity consists of multiple cycles of laser pulses
to manipulate the internal and motional states of the atoms.
Figure \ref{f-setup}A shows the schematic of the energy levels and the cooling sequence
in our setup.
Each cooling cycle starts with the sodium atom in the $|F=2, m_F=-2\rangle$
ground electronic state and a certain vibration state $n$.
In the first step, a Raman pulse drives a transition to the motional state $n-\Delta n$
to reduce the motional energy while also changing the internal state to $|F=1, m_F=-1\rangle$.
In the second step, which finishes the cooling cycle,
an optical pumping pulse bring the atom back to the $|F=2, m_F=-2\rangle$ state to take away
the entropy via the spontaneously emitted photon.
The second step could also change the motional state of the atoms and result in heating.
The possibility for this to happen for an atom in the motional level $n$
is approximately proportional to the effective Lamb-Dicke (LD) parameter
$\eta^{OP}_{\eff}=\sqrt{2n+1}\eta^{OP}$ where $\eta^{OP}=k^{OP}\sqrt{\hbar/2m\omega}$
is the LD parameter for optical pumping (OP), $k^{OP}$ is the wavenumber of the OP light, $m$ is the mass of the atom, and $\hbar$ is the reduced Planck constant.

@@

In order to perform Raman sideband cooling efficiently and minimize
the heating during the optical pumping, we need to have a low initial temperature and
a small LD parameter. However, since the LD parameter $\eta\propto 1/(\sqrt[4]{m}\lambda)$ where $m$ is the mass of the atom and $\lambda$ is the wavelength of the atomic transition,
it is larger for sodium at a given trap depth due to its low mass and short D line wavelength.
With $45$ mW of power in the trap, which is what we use during the cooling sequence,
the optical pumping LD parameters are
$\{\eta^{OP}_1,\eta^{OP}_2,\eta^{OP}_3\} = \{0.602(5), 0.241(2), 0.206(1)\}$
Moreover, due to the small hyperfine splitting in the $3^2P_{3/2}$ manifold,
the sub-Doppler cooling in sodium is also less efficient and our initial temperature of 70 $\mu$K
is higher than what is achieved for other atoms in previous experiments
\cite{Thompson2013,Kaufman2012,Li2012,Liu2017}.
Combined with the high $\eta$, this gives us a initial effective optical pumping LD parameters of
$\{\eta^{OP}_{1\eff},\eta^{OP}_{2\eff},\eta^{OP}_{3\eff}\} = \{4.0(1), 0.67(2), 0.50(1)\}$
As a result, there is a very high average change of motional state during the optical pumping step
(figure \ref{f-ld}B). The averaged change in motional states over the initial distribution
is $2.4$ in the axial direction.
Compare to $0.89-0.95$ in the axis with the weakest confinement in previous experiments
\cite{Thompson2013,Kaufman2012,Li2012,Liu2017} this causes greater uncertainty in the motional
state making it harder to cool.

@@

For the more tightly confined radial directions,
the radial sideband spectra of the initial distribution is shown in figure \ref{f-radial}A
as orange and red lines, where we can clearly see the first order heating,
first order cooling and second order cooling sidebands.
After applying around 1000 cooling pulses cooling in all three dimensions
starting with cooling on the radial second order,
the Raman spectrum with the same parameters is shown in figure \ref{f-radial}A as
blue and green lines, where the first and second order cooling sidebands
on both axis are suppressed.
Given the absence of the second order cooling sideband,
we can estimate the ground state probability in each direction based on the height of
the first order cooling and heating sideband to be $90(2)\%$ and $94(3)\%$.

@@

The axial direction is much less confined which means a much higher initial
effective LD parameter of $\eta^{OP}_{eff1}\approx 4.0(2)$ and is therefore more difficult to cool.
The red line in figure \ref{f-axial}A shows the Raman spectrum in the axial direction
before applying Raman sideband cooling; the clearly resolved sidebands
up to eighth order indicates population in highly-excited motional levels.
Therefore, we begin our cooling sequence by driving Raman transitions on the eighth order axial
cooling sideband. After applying the cooling sequence identical to the one we use to cool
the radial directions, the spectrum is shown in figure \ref{f-axial}A as the blue line.
All of the high order ($\geqslant2$) axial cooling sidebands disappear, and the first order
cooling sideband is strongly suppressed.
The ground state probability calculated from this spectrum is $92(3)\%$.
We can again use Rabi flopping on the carrier and heating sideband to verify this result
similar to the radial direction. In this case, we see very good agreement on the carrier
(figure \ref{f-axial}B) but there is additional decoherence on the axial first order
heating sideband (figure \ref{f-axial}C).
We believe this decoherence is caused by technical noise, which is the most pronounced
on this transition due to its slow Rabi frequency.
The decoherence time scale is consistent with magnetic field fluctuations of $1.5$ mG that we have measured in the lab, which produces
a Zeeman shift of around $3$ kHz.

Combining the axial and radial cooling results,
we have achieved a 3D ground state probability of $77(4)\%$.
We believe this is currently limited by off-resonant scattering from the Raman beams,
which are measured to be between $3$ to $15$ kHz for the detuning and power we use,
and the resonance shift caused by magnetic field fluctuations.
Other imperfections of the current experiment include a $5\%$ loss during the single atom
imaging and an additional $10\%$ due to the heating from the Raman beams during cooling.
We are planing to improve these by increasing the detuning of the Raman beams,
implementing better control of the magnetic field and optimize the temperature during imaging.
We expect these changes to improve the cooling performance even further.
We have shown that despite the difficulty in cooling sodium atoms,
it is possible to achieve reliable three dimensional cooling with significant ground state
probability using new cooling techniques and a optimized cooling sequence.
This opens up the possibility of cooling a large variety of systems to their motional ground
state using Raman sideband cooling that is challenging to laser cool
including molecules and exotic atoms.

% Missing:
% ?? Blackman pulse vs square pulse
